---
title: "Hands-on Exercise 5a"
subtitle: "Creating Ternary Plots with R"
author: "Nazia Faruqui"
date: 2026-02-07
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1. Overview

[Ternary plots](https://en.wikipedia.org/wiki/Ternary_plot) are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) Its display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, we will learn how to build ternary plots programmatically using R for visualising and analysing the population structure of Singapore.

This hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.

-   Derive three new measures using *mutate()* function of **dplyr** package.

-   Build a static ternary plot using *ggtern()* function of **ggtern** package.

-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

# 2. Getting Started

::: panel-tabset
## Installing and loading packages

For this exercise, two main R packages will be used, they are:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

We will also need to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

The code chunk below will be used to accomplish this task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## Importing Data

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. It is called *respopagsex2000to2018_tidy.csv* and is in csv file format. ***read_csv()*** function of **readr** package will be used to import the data into R.

```{r}
pop_data <- read_csv("../data/respopagsex2000to2018_tidy.csv") 
```

## Preparing the data

Next, we will use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```
:::

# 3. Plotting Ternary Diagram with R

## 3.1 Plotting a static ternary diagram

We will use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

## 3.2 Plotting an interactive ternary diagram

The code below creates an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

# 4. Reference

-   Kam, T.S. (2023). [Creating Ternary Plot with R](https://r4va.netlify.app/chap13).
